---
title: "Landmark displacement measurements"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

This vignette illustrates the differences in measuring landmark displacement (or variation).

## Requirements and data

### The packages

This test will be using the `landvR` package from github and will require `devtools`, `geomorph` and `dispRity` from the CRAN.
All these packages can be loaded easily as follows

```{r, message = FALSE, warning = FALSE}
## Loading the libraries (and installing if necessary)
if(!require(devtools)) install.packages("devtools")
if(!require(geomorph)) install.packages("geomorph")
if(!require(landvR)) install_github("TGuillerme/landvR")

## Setting a random seed for the repeatability of this tutorial
set.seed(42)
```

### The data

This whole method is based on Procrustes superimpositions that can be easily done through the `geomorph` package (see the `geomorph` manual and especially the function `gpagen` for more details).
For this example, we will be using the `plethodon` dataset from the `geomorph` `R` package.

```{r}
## Loading the dataset
data(plethodon)

## Generating a Procrustes superimposition
procrustes <- gpagen(plethodon$land, print.progress = FALSE)
```

This results in a dataset of 40 Procrustes superimposition of 12 landmarks each.

# Landmark displacement

First lets get the two most different specimens for simplifying the comparisons:

```{r}
## Getting the two most different specimens
extreme_spec <- variation.range(procrustes, return.ID = TRUE)$min.max

## Separating the specimens
specimen_1 <- procrustes$coords[, , extreme_spec[1]]
specimen_2 <- procrustes$coords[, , extreme_spec[2]]
```

We can then measure the landmark displacements between a focal specimen (specimen 1, in blue below) and a reference one (specimen 2, in orange).
Typically the focal one can be all the observed specimens and the reference one could be the mean or median landmark configuration.

These landmark displacements can be measured using two coordinates systems: the spherical or the vector one and is always measured between $F_{n}$ and $R_{n}$, respectively the $n^{th}$ landmark of the focal specimen and the reference.
$F_{n}$ and $R_{n}$ are defined by two or three coordinates _x, y, z_ depending on the number of dimensions in the system.

## Spherical coordinates

The spherical coordinates measure the landmark's displacement in spherical coordinates systems for each landmark individually.
In other words, the values for each landmark displacement is discrete in space (i.e. independent from other landmarks).

The [spherical coordinate system](https://en.wikipedia.org/wiki/Spherical_coordinate_system) is a 3D (or 2D) referential centred on a point of origin (here the $n^{th}$ landmark of the focal specimen).
The system is defined by two or three variables (for respectively 2D and 3D systems):

 * $\rho$, the Euclidean distance between $F_{n}$ and $R_{n}$.
 * $\phi$, the azimuth angle formed by the projection of $R_{n}$ on the equatorial plan ($f_{(x)}=0$).
 * $\theta$ (for 3D only), the polar angle formed by the projection $R_{n}$ on the polar plan ($f{(y)=0}$)

In `landvR` they are measured using basic algebra and trigonometry through the following function:

```{r}
## Measuring the spherical coordinates between specimen 1 (focal) and 2 (reference)
spherical_diff <- coordinates.difference(coordinates = specimen_1, reference = specimen_2,
                                        type = "spherical", angle = "degree")[[1]]
```

This results in a list of coordinates for each specimen.
For example, to look at the differences between the first specimen and the consensus in terms of spherical coordinates, one can call:

```{r}
## The differences between the two selected specimens
spherical_diff
```

In this case, for each landmark, the longer the radii $\rho$, the more it is displaced and the bigger the azimuth $\phi$, the more away from the equatorial plan.

Here is a graphical representation of the system for a single landmark (3) with a radius of `r spherical_diff[3, "radius"]` and an angle of `r spherical_diff[3, "azimuth"]` degrees.


```{r, fig.height = 8, fig.width = 8, echo = FALSE}
## Graphical parameters
par(mar = c(0, 0, 0, 0))

## The variation range between the two specimens
procrustes.var.plot(specimen_1, specimen_2, col = list("blue", "white"))
legend(0.4, 0.4, legend = c("Focal specimen", "Reference spec."), pch = 19, col = c("blue", "orange"),
      bty = "n")
points(specimen_2, col = "orange", pch = 19)

## Getting the spherical coordinates representation for the landmark 3
landmark <- 3

## Spherical coordinate system
circle_center <- specimen_1[landmark,]
circle_radius <- spherical_diff[landmark, "radius"]
azimuth_angle <- spherical_diff[landmark, "azimuth"]
plotrix::draw.circle(x = circle_center[1], y = circle_center[2], radius = circle_radius)

## radius
arrows(x0 = specimen_1[landmark, 1], y0 = specimen_1[landmark, 2],
       x1 = specimen_2[landmark, 1], y1 = specimen_2[landmark, 2],
       length = 0, angle = 0, code = 2, lwd = 1.5)

## azimuth
arrows(x0 = specimen_2[landmark, 1], y0 = specimen_1[landmark, 2],
       x1 = specimen_1[landmark, 1] + abs(specimen_2[landmark, 1] - specimen_1[landmark, 1]),
       y1 = specimen_1[landmark, 2],
       length = 0, angle = 0, code = 2, lwd = 1, lty = 3)
    
## angle
plotrix::draw.arc(specimen_1[landmark, 1], specimen_1[landmark, 2], 0.04, deg1 = 180, deg2 = azimuth_angle + 180)

## Explanatory arrows

## The circle:
circle_text_pos <- c(0.1, 0.2)
arrows(x1 = 0, y1 = 0.075, x0 = circle_text_pos[1], y0 = circle_text_pos[2], length = 0.1)
text(x = circle_text_pos[1], y = circle_text_pos[2] + 0.01, "Polar coordinates system")

## The radius
rad_pos <- -0.2
arrows(x1 = mean(c(specimen_1[landmark, 1], specimen_2[landmark, 1])),
       y1 = mean(c(specimen_1[landmark, 2], specimen_2[landmark, 2])) - 0.01,
       x0 = 0, y0 = rad_pos, length = 0.1)
text(0, rad_pos-0.02, "radius")

## The azimuth
rad_pos <- -0.2
arrows(x1 = specimen_1[landmark, 1] + abs(specimen_2[landmark, 1] - specimen_1[landmark, 1]),
       y1 = specimen_1[landmark, 2],
       x0 = 0.4, y0 = rad_pos, length = 0.1)
text(0.4, rad_pos-0.02, "azimuth")


## The angle
rad_pos <- -0.2
arrows(x1 = specimen_2[landmark, 1]+0.02, y1 = specimen_1[landmark, 2]-0.01,
       x0 = rad_pos + 0.05, y0 = 0.03, length = 0.1)
text(rad_pos, 0.03, "angle from\nazimuth")

```

This system has the advantage of directly measuring the landmark displacement in _any_ direction and to be independent from other landmarks within the same specimen.
However, the azimuth and polar angle are often hard to interpret in isolation since they are relative to the equatorial and polar plan of the specimens (which can be arbitrary).


## Vector coordinates

The vector coordinates system is a bit different than the spherical one in that it is based on the landmark displacements relative to the centroid of the reference specimen.
The centroid point of the reference, $C$, is the mean of all the _x, y, z_ coordinates for all the landmarks of the reference specmin.
This coordinate system is based on the vectors between the centroid $C$ and $F_{n}$ or $R_{n}$.
These two vectors are called $\bf{v}_{Fn}$ and $\bf{v}_{Rn}$ respectively.
The system is defined by two variables (regardless of the number of dimensions):

 * the displacement length which is the difference between both vectors ($||\bf{v}_{Rn}|| - ||\bf{v}_{Fn}||$)
 * the angle between the two vectors $\bf{v}_{Fn}$ and $\bf{v}_{Rn}$

In `landvR` they are measured using basic vector operations through the following function:

```{r}
## Measuring the spherical coordinates between specimen 1 (focal) and 2 (reference)
vector_diff <- coordinates.difference(coordinates = specimen_1, reference = specimen_2,
                                      type = "vector", angle = "degree",
                                      absolute.distance = FALSE)[[1]]
```

Note that we've asked the distances to be not absolute, toggling this option allows the distances to be calculated as $\sqrt((\bf{v}_{Fn} - \bf{v}_{Rn})^2)$ (`TRUE`) or \bf{v}_{Fn} - \bf{v}_{Rn} (`FALSE`).
With absolute distances, it measures only the _magnitude_ of the displacement whereas without absolute distances the results also indicate the _direction_ of the displacement: a positive value means the landmark is displaced "outside" of the reference and a negative value means it is displaced "inside".


```{r}
## The differences between the two selected specimens
vector_diff
```

In this case, for each landmark, the longer the difference in length (in absolute value), the more it is displaced and the bigger the angle, the more it is offset.
In terms of biology, if all lengths are positive and all angles are close to 0, the specimen is simply generally bigger (with no major changes in shape).
On the other hand, if the lengths are close to 0 but the angles are high, the specimen is not bigger but still "deformed",

Here is a graphical representation of the system for a single landmark (7) with a difference of `r vector_diff[7, "length"]` and an angle of `r vector_diff[7, "angle"]` degrees:

```{r, fig.height = 8, fig.width = 8, echo = FALSE}
## Graphical parameters
par(mar = c(0, 0, 0, 0))

## The variation range between the two specimens
procrustes.var.plot(specimen_1, specimen_2, col = list("blue", "white"))
legend(0.4, 0.4, legend = c("Focal specimen", "Reference spec."), pch = 19, col = c("blue", "orange"),
      bty = "n")
# legend(0.4, 0.4, legend = c("Focal specimen", "Reference spec.", "Ref. centroid"),
#    pch = c(19, 19, 13), col = c("blue", "orange", "black"), bty = "n")
points(specimen_2, col = "orange", pch = 19)

## Add the centroid
ref_centroid <- apply(specimen_2, 2, mean)
points(ref_centroid[1], ref_centroid[2], pch = 13, cex = 2)

## Getting the spherical coordinates representation for the landmark 7
landmark <- 7

## Get the length and angle
vector_length <- vector_diff[landmark, "length"]
vector_angle <- vector_diff[landmark, "angle"]
## Get the angle

## Ref dist
arrows(x0 = ref_centroid[1], y0 = ref_centroid[2],
       x1 = specimen_2[landmark, 1], y1 = specimen_2[landmark, 2],
       length = 0, lty = 3)

## Focal dist
arrows(x0 = ref_centroid[1], y0 = ref_centroid[2],
       x1 = specimen_1[landmark, 1], y1 = specimen_1[landmark, 2],
       length = 0, lty = 3)

## Angle
fun.angle <- function(one_row, axis, dimension) {
    return(
        acos( ( sqrt((one_row[-c(1:dimension)][axis] - one_row[1:dimension][axis])^2) ) / sqrt(sum((one_row[-c(1:dimension)]-one_row[1:dimension])^2)) )
        )
}
angle_shift <- fun.angle(c(c(0,0), specimen_2[landmark, ]), 2, 2)*180/pi+90
plotrix::draw.arc(ref_centroid[1], ref_centroid[2], 0.2, deg1 = angle_shift, deg2 = vector_angle + angle_shift)

## Distance projection
projection <- dist(rbind(ref_centroid, specimen_2[landmark,]))[[1]]
plotrix::draw.arc(ref_centroid[1], ref_centroid[2], projection, deg1 = angle_shift, deg2 = vector_angle + angle_shift, lty = 3)

## Get the projected x/y coords
x_proj = projection * cos(vector_angle+angle_shift-90)
y_proj = -projection * sin(vector_angle+angle_shift-90) # Should be around 0.0265!

## The measured distance
arrows(x0 = x_proj, y0 = 0.0265,
       x1 = specimen_1[landmark, 1], y1 = specimen_1[landmark, 2],
       length = 0, lty = 1, lwd = 1)

## Explanatory arrows

## The centroid:
text_pos <- c(0.2, -0.2)
arrows(x1 = ref_centroid[1]+0.02, y1 = ref_centroid[2]-0.02,
       x0 = text_pos[1], y0 = text_pos[2], length = 0.1)
text(x = text_pos[1], y = text_pos[2]-0.015, "Reference's centroid")

## The angle
text_pos <- c(-0.1, -0.2)
arrows(x1 = mean(c(specimen_1[landmark, 1], ref_centroid[1])) - 0.04,
       y1 = mean(c(specimen_1[landmark, 2], ref_centroid[2])),
       x0 = text_pos[1], y0 = text_pos[2], length = 0.1)
text(x = text_pos[1], y = text_pos[2]-0.015, "angle")

## The length
text_pos <- c(-0.3, -0.2)
arrows(x1 = specimen_1[landmark, 1] + 0.015,
       y1 = specimen_1[landmark, 2] - 0.005,
       x0 = text_pos[1], y0 = text_pos[2], length = 0.1)
text(x = text_pos[1], y = text_pos[2]-0.03, "length\nv(ref) - v(foc)")


## The reference vector
text_pos <- c(-0.2, 0.3)
arrows(x1 = specimen_2[landmark, 1] + 0.03,
       y1 = specimen_2[landmark, 2] - 0.005,
       x0 = text_pos[1], y0 = text_pos[2], length = 0.1)
text(x = text_pos[1], y = text_pos[2]+0.02, "reference vector (v(ref))")


## The length
text_pos <- c(0.2, 0.2)
arrows(x1 = ref_centroid[1] - 0.1,
       y1 = ref_centroid[2] + 0.015,
       x0 = text_pos[1], y0 = text_pos[2]-0.02, length = 0.1)
text(x = text_pos[1], y = text_pos[2], "focal vector (v(foc))")

```

This coordinate system has the advantage to be more interpretable biologically (i.e. the length and angle will give a clear overview of how the landmark has moved relative to the reference) but the measurements are always relative to the reference and dependent on the other landmarks (used to calculate the centroid point).

